DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-08-03 2015CompleteDerivation/completeDerivation.tex
Reviewed classical assumption with Christine

	The first problem is methodological. To be able to talk about the states of a part we need a physical process that is able to distinguish between them. For a pool ball we can imagine marking one spot with a red marker. This allows us to track it as the ball moves or collides with other balls. For an electron, instead, we do not have a way to mark a piece. In fact when two electrons scatter we can't even tell which is which, let alone what portion went where. The classical assumption may not hold because we do not have suitable physical processes at our disposal.

@@

	%The other idealization is when the interaction between system and environment is not symmetrical. For example, suppose the electric charge of the system under study is much larger than the charges in the environment. The electromagnetic field generated can be considered determined only by our system with good approximation. But this also breaks down: if each part of the environment is made of the same homogeneous material the charge density will be the same and the interaction between particles will be symmetrical.

	Another issues is that defining a state requires some kind of asymmetry between system under study and environment. The system evolution is not affected by the state of the environment, so that we can define an independent state, yet the environment evolution is affected by the state of the system, so that we have external processes that allow us to define physical distinguishability. If we try to assume that both the system and environment are really made of the same classical material, then the claim to that asymmetry is lost: there is no difference in size, everything is made of infinitesimal pieces, and there is no difference in the laws of evolution, as the pieces are homogeneous.

	Another issue arises if we include the probe in our description. Say $a \in A$ is the state of our system and $b \in B$ the state of our probe, the future state $\hat{a}=f(a)$ because the system is deterministic, but $\hat{b}=g(a,b)$ because the probe needs to be affected by the system. This means $(a, b) \mapsto (\hat{a}, \hat{b})$ is not a reversible map. If we want determinisms and reversibility for the combined system, we either have two isolated systems $\hat{a}=f(a)$ and $\hat{b}=g(b)$, which means the probe can't learn anything about the system, or two interacting systems $\hat{a}=f(a, b)$ and $\hat{b}=g(a, b)$, which means our system is not deterministic as its future state is determined by the state of the probe. In other words: we can't expect to have deterministic and reversible evolution at all levels of aggregations and have the parts interact in any physically meaningful way.

	%Therefore our justification for holding the assumption has to be that the pieces are small compared to some scale (they can be regarded as infinitesimal) but large compared to another (they are little affected by the environment). For example, water is a homogeneous material if we consider it made up of parts sufficiently small for the scale of the problem at hand but big enough to contain a large number of molecules such that the average motion is sufficiently stable.

	%Other conceptual problems stem from the following two competing requirements: on one side we want the system under study to be isolated so that we can define independent states; on the other side we want the environment to not be isolated so that we can define physical processes that interact with the system. The problem is that the same physical object could be either under study or part of the environment depending on the situation. In other words: we want systems to be interacting so that they can be physically aware of each other, yet we want them not interacting so that we can properly define states on them. We clearly can't have it both ways.

	The same problem of not meaningful interaction surfaces when we consider the states of infinitesimal parts. Under our classical assumption, if we were to take the state of an infinitesimal part to really be its full description, with no unstated part, then each piece would evolve independently, oblivious to the other parts. Each particle would essentially reside in its own separate physical universe, as it cannot physically distinguish anything else. This is not physically meaningful.

	If we assumed the system is deterministic and reversible only as a whole, then each part could evolve depending on the states of the other parts. This would seem much better, as the state of the pieces is still exhaustive yet they are physically connected to each others. But this does not actually solve the problem of independent oblivious components. First, for this to work, one would have to specify how the states of the parts were defined since their evolution is no longer deterministic (the future of each part depends on the state of other parts). One would also need to explain what type of interaction exists between the parts: simply using conservative forces, for example, would smell of a circular argument since those are precisely the force that we derive in the case where the pieces do evolve independently. But ignoring these two issue, the bigger problem is that we can always locally separate the evolution into independent degrees of freedom. For example, the position and momentum of two particles may affect each other during the evolution, but the average and difference in position and momentum may evolve independently.\footnote{Also note that any Hamiltonian is locally isomorphic to a free particle (citation needed)} We can always find such local decomposition, and the easiest way to see that is in terms of the transported variables: they retain the original value, they clearly evolve independently. And since the pieces are infinitesimal, a local decomposition is all that is needed. While such description may be cumbersome to achieve in practice, conceptually it is still possible. So, even if each particle evolution depends on the state of the others, the system is described by degrees of freedom that evolve independently, oblivious to each other.

	The moral is that the classical idea of being able to assign to all systems a state which represents its full description does not work. The state is only what we can describe and it can't be the full description. As the division between system and environment is subjective, each system must be able to function as both. Therefore it will have a part whose evolution depends only on the system itself, the state, and a part whole evolution depends on other systems, the unstated part. While they may not be the same in all circumstances, they must exist. The state is what gives the identity to the system, its the subject of our discussion. The unstated part is what allows continuous interaction between the system and the environment, its what allows us to have the discussion. As we'll see later, it is precisely this problem that quantum system addresses better conceptually than classical mechanics.

	One final problem is with time itself: an infinitesimally reducible deterministic and reversible system cannot tell time. If it did, we would be able to find a quantity that changes through time but is invariant under state variable changes. The problem is that deterministic and reversible time evolution is equivalent to a state variable change. In fact, at a fixed time, we can choose to describe the system with the transported state variables of any past or future times. So, all quantities that are invariant under state variable changes are also invariant under time evolution.

@@

\section{Elementary systems}

@@

The link between kinematics, the study of motion through quantities like velocity and acceleration, and dynamics, the study of laws of evolution, is crucial in physics. In Newtonian physics, $\vec{F}=m\vec{a}$ provides such link. In Hamiltonian and Lagrangian mechanics, the link is implicitly given by the choice of the Hamiltonian and the Lagrangian.

In this work we make the link conceptually explicit by assuming the following:

\begin{assump}[Kinematic equivalence]\label{kinematic_equivalence}
	Studying the kinematic of the system (i.e. trajectories in space) is equivalent to study the dynamics of the system (i.e. state trajectories in state space).
\end{assump}

@@

>>>>

	The first problem is methodological. To be able to talk about the states of a part we need a physical process that is able to distinguish between them. For a billiard ball we can imagine marking one spot with a red marker. This allows us to track it as the ball moves or collides with other balls. For an electron, instead, we do not have a way to mark a piece. In fact when two electrons scatter we can't even tell which is which, let alone what portion went where. The classical assumption may not hold because we do not have suitable physical processes at our disposal.

@@

	Another issue is that defining a state requires some kind of asymmetry between system under study and environment. The future state of the system does not depend on the state of the environment, so that we can define an independent state, yet the future state of the environment is affected by the state of the system, so that we can have external processes that allow us to define physical distinguishability. If we try to assume that both the system and environment are really made of the same classical material, then the claim to that asymmetry is lost: there is no difference in size, everything is made of infinitesimal pieces, and there is no difference in the laws of evolution, as the pieces are homogeneous.

	Another issue arises if we include the probe in our description. Say $a \in A$ is the state of our system and $b \in B$ the state of our probe, the future state $\hat{a}=f(a)$ because the system is deterministic, but $\hat{b}=g(a,b)$ because the probe needs to be affected by the system. This means $(a, b) \mapsto (\hat{a}, \hat{b})$ is not a reversible map. If we want determinisms and reversibility for the combined system, we either have two isolated systems $\hat{a}=f(a)$ and $\hat{b}=g(b)$, which means the probe can't learn anything about the system, or two interacting systems $\hat{a}=f(a, b)$ and $\hat{b}=g(a, b)$, which means our system is not deterministic as its future state is determined by the state of the probe. In other words: we can't expect to have deterministic and reversible evolution at all levels of aggregation and have the parts interact in any physically meaningful way.

	The same problem of physically meaningless interaction surfaces when we consider the states of infinitesimal parts. Under our classical assumption, if we were to take the state of an infinitesimal part to really be its full description, with no unstated part, then each piece would evolve independently, oblivious to the other parts. Each particle would essentially reside in its own separate physical universe, as it cannot physically distinguish anything else. This is not physically meaningful.

	If we assumed the system is deterministic and reversible only as a whole, then each part could evolve depending on the states of the other parts. This would seem much better, as the state of the pieces is still exhaustive yet they are physically connected to each other. But this does not actually solve the problem of independent oblivious components. First, for this to work, one would have to specify how the states of the parts were defined since their evolution is no longer deterministic (the future of each part depends on the state of other parts). One would also need to explain what type of interaction exists between the parts: simply using conservative forces, for example, would hint at a circular argument since those are precisely the forces that we derive in the case where the pieces do evolve independently. But ignoring these two issues, the bigger problem is that we can always locally separate the evolution into independent degrees of freedom. For example, the position and momentum of two particles may affect each other during the evolution, but the average and difference in position and momentum may evolve independently.\footnote{Also note that any Hamiltonian is locally isomorphic to a free particle (citation needed)} We can always find such local decomposition, and the easiest way to see that is in terms of the transported variables: they retain the original value, they clearly evolve independently. And since the pieces are infinitesimal, a local decomposition is all that is needed. While such description may be cumbersome to achieve in practice, conceptually it is still possible. So, even if each particle evolution depends on the state of the others, the system is described by degrees of freedom that evolve independently, oblivious to each other.

	The moral is that the classical idea of being able to assign to all systems a state which represents their full description does not work. The state is only what we can describe and it can't be the full description. As the division between system and environment is subjective, each system must be able to function as both. Therefore it will have a part whose evolution depends only on the system itself, the state, and a part whose evolution depends on other systems, the unstated part. While they may not be the same in all circumstances, they must exist. The state is what gives the identity to the system, the part we can study and describe. The unstated part is what allows continuous interaction between the system and the environment, it's what allows us to study and describe the state. As we'll see later, it is precisely this problem that quantum mechanics addresses better conceptually than classical mechanics.

	One final problem is with time itself: an infinitesimally reducible system undergoing deterministic and reversible evolution cannot tell time. If it did, we would be able to find a quantity that changes through time but is invariant under state variable changes. The problem is that deterministic and reversible time evolution is equivalent to a state variable change. In fact, at a fixed time, we can choose to describe the system with the transported state variables of any past or future times. So, all quantities that are invariant under state variable changes are also invariant under time evolution. As the very definition of deterministic and reversible processes relies on time, this has to come from elsewhere (i.e. the environment).

@@

\section{Kinematics and elementary systems}

@@

The link between kinematics, the study of motion through quantities like velocity and acceleration, and dynamics, the study of laws of evolution, is crucial in physics. In Newtonian physics, $\vec{F}=m\vec{a}$ provides such link. In Hamiltonian (and Lagrangian) mechanics, the link is implicitly given by the choice of the Hamiltonian (or Lagrangian).

Here we make the link conceptually explicit by assuming the following:

\begin{assump}[Kinematic equivalence]\label{kinematic_equivalence}
	For the given system, studying its kinematics  (i.e. trajectories in space) is equivalent to study its dynamics  (i.e. state trajectories in state space).
\end{assump}

@@

\subsection{Space-time manifold and Lagrangian mechanics}

As we need to describe trajectory, we introduce $\mathcal{M}$ as the manifold that defines space and time. We call $x^i$ the spatial coordinates and $t$ the time coordinate. A trajectory will be given by a set of functions $x^i(s)$ and $t(s)$. What we need to understand is the link between these trajectories and the complete state space of the particles of a classical material.

We first note that the system of units for the trajectories is fully defined by $\mathcal{M}$. Any trajectory can be specified once a set of coordinates $(t, x^i)$ is chosen. Under the kinematic assumption, these units must also be sufficient to define the dynamics. Therefore $\mathcal{M}$ has conceptually two functions: it is the space where the motion unfolds and it is the space that defines the system of units for our states. This already tells us that the our state space is a hyper-surface of $T^*\mathcal{M}$.

Under the kinematic equivalence, given a trajectory we must identify a state and vice-versa. There must be a homeomorphism between the complete state space and the space of possible trajectories. We already know what the complete state space is. We just need to find a way to label the trajectories so that we can express the correspondence.

Because of the double role of $\mathcal{M}$, we can set half of our state variables to be equal to the space-time variables: $(t, q^i) = (t, x^i)$. This already tells us that the trajectory is differentiable since $q^i(t)$ is differentiable. The four velocity $(u^t, u^i) = (d_s t, d_s x^i)$ is therefore well defined and it can be used to identify trajectories. The magnitude of the four velocity can be rescaled by changing the evolution parameter $s$, and is therefore not physical. Only the direction is physical so the four velocity gives us another $n$ independent variables to identify trajectories. The combined position and velocity, which we call \emph{initial conditions}, identify $2n + 1$ trajectories using a hyper-surface of the tangent bundle $T\mathcal{M}$. This is exactly the dimensionality of the complete state space $\bar{\mathcal{S}} \subset T^*\mathcal{M}$. We cannot add more curves, as a homeomorphism preserve dimensionality, therefore we found the suitable space for our trajectories.

As $\omega(dt,dq^i,dE,dp_i)$ allows us to count possibilities in terms of state variable, we will have a metric $g(dt,dx^i,du^t,du^i)$ that allows us to count the possibilities in terms of initial conditions. Since the both $(dt,dx^i)$ and $(du^t,du^i)$ are four vectors and the metric must be linear, $g$ is a rank two tensor on $\mathcal{M}$. $g$ is also symmetric. Suppose $(t, x^i)$ are orthogonal directions, then a change of unit of one coordinate will only change the component of position and velocity for that coordinate. That is: the component of position and velocity form an independent d.o.f.. For that coordinate system it must be $g(dx^i, du^j)=0=g(dx^j, du^i)$ for $i\neq j$. And the same must hold across temporal and standard d.o.f.. As $g$ is a tensor, the symmetry applies to all choices of coordinates.

Mathematically, under the kinematic assumption the state space $\bar{\mathcal{S}} \subset T^*\mathcal{M}$ is isomorphic to a hyper-surface of the tangent bundle $T\mathcal{M}$ of a pseudo-Riemannian manifold $(\mathcal{M}, g)$.

\begin{prop}\label{riemannian_manifold}
	The state space $\bar{\mathcal{S}} \subset T^*\mathcal{M}$ is isomorphic to a hyper-surface of the tangent bundle $T\mathcal{M}$ of a pseudo-Riemannian manifold $(\mathcal{M}, g)$.
\end{prop}

\begin{justification}
	TODO
\end{justification}

\subsection{Scalar and vector potential forces}
